---
title: "Question 3"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r include=FALSE}
#### 0. IMPORT LIBRARIES & DATA ####
library(likert)
library(reshape2) # contains melt function for converting tables to charts
library(magrittr)
library(haven)
library(ggplot2)
library(dplyr)
library(gridExtra)

d <- read_dta("data/anes_timeseries_2020_stata_20210211.dta")

```
# 1. Introduction

### Does a case of COVID-19 in someone's household change how they approve of their governor's response to the pandemic?
Before the advent of COVID-19, America's social fabric was already stretched thin by accelerating political polarization. Now, with the advent of COVID-19 and the damage it has wrought, that fabric appears at risk of being torn asunder. COVID-19 has become the gravitational center around which all other contemporary American politics revolve. 

The federal government has left much of the pandemic response to state and local governments. A scattered and inconsistent landscape of policies that has gradually taken shape implies that our personal experiences of the pandemic have been highly regionally based. Yet with the country as divided as it is - a division which also has geographical components - does experiencing COVID-19 up close and personal impact people's perception of their local official's handling of the pandemic? Or do the tribes stick to their pre-baked narratives regardless of experience? The answer to this question could have implications for further efforts to de-radicalize the ever growing political division in our country.

In this report, we ask: are people who have shared a house with someone with more likely to approve or disapprove of their governor's pandemic response?

```{r include=FALSE}
#### 1. CLEAN DATA ####
# NOTE: See code book on ANES' website for more detail on these codes.
d_trim = d[, c('V201018', 'V201507x', 'V201151', 'V201153',
               'V201624', 'V201625', 'V201066', 'V201145', 'V201146')]

# Rename columns
colnames(d_trim) = c('party_of_reg', 'age', 'sentiment_biden', 'sentiment_harris',
                     'covid_test_positive', 'covid_symptoms',
                     'vote_for_governor', 'gov_approval', 'gov_how_much')

# Trim columns relevant to q3
q3 <- d_trim[, c('covid_test_positive', 'covid_symptoms',
               'vote_for_governor', 'gov_approval', 'gov_how_much')]

# Clean the labels from the spss dataset
q3_clean <- zap_labels(q3)
```

```{r include=FALSE}
#### 2. PROCESS DATA - CREATE NEW COLUMN WITH LIKERT SCALE####
# Define function that will rank, from 1 to 5 on the likert scale, based on the two columns in which approval data is currently located (one is just approve/disapprove, the next is strongly or not).This function will then be applied to those column vectors, and a new column with a 1-5 likert score created, which will be used for the Rank-sum hypothesis test and visualizations.

# (1,1) = strongly approve, (1,2) = approve, -8 = neutral, 
# (2,2) = disapprove, (2,1) = disapprove strongly
scale_approval <- function(approve_disapprove, how_much){
  if (approve_disapprove == 1 & how_much == 1) {
    # "strongly approve"
    return(1)
  } else if (approve_disapprove == 1 & how_much == 2) {
    # "approve"
    return(2)
  } else if (approve_disapprove == -8){
    # "neutral"
    return(3)
  } else if (approve_disapprove == 2 & how_much == 2){
    # "disapprove"
    return(4)
  } else if (approve_disapprove == 2 & how_much == 1) {
    # "strongly disapprove"
    return(5)
  } else {
    return(-88) # needs this for cases where some version of the above isn't met
  }
}



# NOTE: -88 is returned in the function above for easy subsetting and filtering below based on positive/negative values.
q3_clean$gov_scale <- mapply(scale_approval, approve_disapprove = q3_clean$gov_approval,
       how_much = q3_clean$gov_how_much)
q3_clean <- subset(q3_clean, gov_scale > 0) # remove all entries with gov_scale == -88

#### NOTES ON CLEANED DATASET AFTER ADDING NEW COLUMN
# There are 46 entries that did not fit into one of the 5 conditional statements
# according to my function scale_approval. These were labeled by the function as
# -88. Upon investigation of these, all fall into one of the following categories:
# 1) R refused to provide an answer for either "approve/disapprove" or "how much"
#    If someone answered "don't know" in "approve/disapprove", they are labeled
#    as neutral. If a respondent answered approve/disapprove but then refused
#    the follow up question about "how much," then that sample is lacking the
#    necessary data to fit into the new column
# 2) Refused to respond at all, resulting in a -9 entry for approve/disapprove
# 
# Since the number of samples that fit into one of the above two categories are
# small, compared to the total sample size, and they lack the data needed to calculate
# the new 1-5 likert valuewhich is needed to run Wilcox.test, I decided to drop them. 
```

### Data: sources & summary
To answer this question, we use data released just last month by the American National Election Studies. This dataset is the preliminary release of the ANES' 2020 pre-election survey, and was collected in collaboration with Westat Inc. and Marketing Systems Group, survey research firms. Resondents were surveyed from August 18, 2020 to November 3, 2020.

The ANES' dataset provides two measures of COVID-19 household prevalance. First, respondents are asked if anyone in their household has tested positive for COVID-19. Second, they are asked if anyone in their household has COVID-19 like symptoms (but not necessarily a positive test result). We operationalize on both of these, and investigate whether people's approval varied significantly in houses with COVID-19 and those without.


```{r echo=FALSE}
num_positive_tests <- as.numeric(count(subset(q3_clean, covid_test_positive == 1)))
num_no_positive_tests <- as.numeric(count(subset(q3_clean, covid_test_positive == 2)))
total_tests <- num_no_positive_tests + num_positive_tests

num_positive_symptoms <- as.numeric(count(subset(q3_clean, covid_symptoms == 1)))
num_no_symptoms <- as.numeric(count(subset(q3_clean, covid_symptoms == 2)))
total_symptoms <- num_positive_symptoms + num_no_symptoms

tab <- as.table(matrix(c(num_positive_tests, num_no_positive_tests, total_tests, 
                         num_positive_symptoms, num_no_symptoms, total_symptoms),
                       nrow=2, byrow=TRUE))

dimnames(tab) <- list(Measure = c("Test", "Symptoms"), 
                      Positive = c("Yes", "No", "Total"))
tab
#sprintf("Households with positive COVID-19 tests: %s", num_positive_tests, quote=FALSE)
#sprintf('Households without a positive COVID-19 test: %s', num_no_positive_tests)
#cat("\n")
#sprintf('Households with COVID-19 symptoms: %s', num_positive_symptoms)
#sprintf('Households without COVID-19 symptoms: %s', num_no_symptoms)



```
**Note:** "total" values are different, since some respondents may have responded to one question about symptoms, but not the other, and vice versa.

Out of a total of 8,144 respondents, 286 (~3.5%) lived in households where someone tested positive for COVID-19. Three and a half times as many people reported as living with someone that exhibited signs of COVID-19 - 993 respondents (~12%)

The ANES' survey includes two questions that gauge respondents' approval or disapproval of their governor's response to the pandemic. The first has three options: approve, disapprove, or "don't know". The second is a binary choice: strongly or not strongly. I combined these two columns into a single five-point Likert scale as follows:
1 - Strongly Approve, 2 - Approve, 3 - Neutral ("don't know"), 4 - Disapprove, 5 - Strongly Disapprove.




```{r echo=FALSE}
# Drop samples where covid test is below 0, then factor gov_scale column
q3_clean_noneg <- subset(q3_clean, covid_test_positive > 0 & 
                           covid_symptoms > 0)

# Recast gov_scale & test_positive columns as factors, and
# recode gov_scale factors from 1-5 to "strongly approve" etc,
# so that they show up in the visualization correctly
q3_clean_noneg$gov_scale <- as.factor(q3_clean_noneg$gov_scale)
q3_clean_noneg$covid_test_positive <- as.factor(q3_clean_noneg$covid_test_positive)

# Logic behind recasting columns with numeric factors to character factors: the likert package provides some nice,
# turnkey graphics for likert-type data. But customizing the plot functionality is a little convoluded, so it's easier
# to just rename/recast the data being fed into the likert plot function.

# This must happen here, rather than above, since the Wilcox.test hypothesis works but calculating numeric
# rank values, then adding them together. If these numeric types are first converted to factors, Wilcox.test fails
q3_clean_noneg$gov_scale <- as.factor(recode(q3_clean_noneg$gov_scale, `1` = "Strongly Approve", `2` = "Approve", 
                                             `3` = "Neutral", `4` = "Disapprove", `5` ="Strongly Disapprove"))

q3_clean_noneg$covid_test_positive <- as.factor(recode(q3_clean_noneg$covid_test_positive, `1` = "Positive test/symptoms",
                                                       `2` = "No positive test/symptoms"))

# q3_clean_noneg$covid_symptoms <- as.factor(recode(q3_clean_noneg$covid_symptoms, `1` = "Person in household had COVID # symptoms", `2` = "No symptoms in household"))

# Table 1
covid_test_positive_table <- table(q3_clean_noneg$gov_scale, q3_clean_noneg$covid_test_positive)
covid_test_positive_table <- round(prop.table(covid_test_positive_table, 2), 2)
covid_test_positive_table_melt <- melt(covid_test_positive_table)
colnames(covid_test_positive_table_melt) <- c("gov_scale", 
                                              "positive_test",
                                              "percent")
# Reference: https://stackoverflow.com/questions/22305023/how-to-get-a-barplot-with-several-variables-side-by-side-grouped-by-a-factor

# https://stackoverflow.com/questions/14622421/how-to-change-legend-title-in-ggplot
two_bar_test <- ggplot(covid_test_positive_table_melt, 
       aes(x=gov_scale, y=percent, fill=factor(positive_test))) +
  geom_bar(stat="identity", position="dodge") +
  xlab("Approval Rating") +
  ylab("%") +
  ggtitle("Do you approve of your governor's COVID-19 response?", subtitle="Tests") +
  scale_fill_discrete("") +
  scale_x_discrete(labels = NULL) + labs(x = "") +
  theme( 
    plot.title = element_text(family = "sans", size = 18, margin=margin(0,0,30,0)),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    legend.position="top") 

# Table 2
covid_symptoms_table <- table(q3_clean_noneg$gov_scale, q3_clean_noneg$covid_symptoms)
covid_symptoms_table <- round(prop.table(covid_symptoms_table, 2), 2)
covid_symptoms_table_melt <- melt(covid_symptoms_table)
colnames(covid_symptoms_table_melt) <- c("gov_scale", 
                                         "symptoms",
                                         "percent")

two_bar_symptoms <- ggplot(covid_symptoms_table_melt, 
       aes(x=gov_scale, y=percent, fill=factor(symptoms))) +
  geom_bar(stat="identity", position="dodge") +
  xlab("Approval rating") +
  ylab("%") +
  ggtitle(label="", subtitle="Symptoms") +
  theme(legend.position="none")



grid.arrange(two_bar_test, two_bar_symptoms)

```

[Comment on low number of neutrals]


```{r}
#### LIKERT PLOTS ####

# This line is necessary since the "likert" function takes a df as input.
# The original error produced when I tried to pass q3_clean_noneg$gov_scale:
# "Error in likert(q3_clean_noneg$gov_scale) : 
# The items parameter must be a data frame. If trying to subset 
# a data frame to analyze only one column, try: items=mydf[,1, drop=FALSE]." 
itemsq3 <- as.data.frame(q3_clean_noneg[,'gov_scale', drop=FALSE])

lgrq3 <- likert(itemsq3, grouping=q3_clean_noneg$covid_test_positive)

plot(lgrq3) + 
  ggtitle("People living with a family member that has tested positive for COVID-19 
  tend to disapprove of their governor's pandemic response")


#### DISTRIBUTION OF GOV_SCALE BAR CHARTS ####

# Subset q3_clean to create two vectors. Wilcox.test two numeric vectors x, y as input. 
samples_positive_covid_test <- subset(q3_clean, covid_test_positive == 1)
samples_negative_covid_test <- subset(q3_clean, covid_test_positive == 2)

# Same logic as above, except for covid_symptoms instead of positive_covid_test.
samples_covid_symptoms <- subset(q3_clean, covid_symptoms == 1)
samples_no_covid_symptoms <- subset(q3_clean, covid_symptoms == 2)

all_bar <- ggplot(data=q3_clean_noneg, mapping=aes(y=gov_scale)) +
  geom_bar() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("All")

# samples_positive_covid_test - samples with a positive covid test
positive_bar <- ggplot(data=samples_positive_covid_test, mapping=aes(y=gov_scale)) +
  geom_bar() +
  coord_flip() +
  scale_y_continuous(breaks=1:5,
                     labels=c("Strongly Approve","Approve",
                              "Neutral","Disapprove","Strongly Disapprove")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Positive Covid test")

# samples_negative_covid_test - samples without a positive covid test
no_positive_bar <- ggplot(data=samples_negative_covid_test, mapping=aes(y=gov_scale)) +
  geom_bar() +
  coord_flip() +
  scale_y_continuous(breaks=1:5,
                     labels=c("Strongly Approve","Approve",
                              "Neutral","Disapprove","Strongly Disapprove")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("No positive Covid test")



# grid.arrange(all_bar, positive_bar, no_positive_bar, ncol=3)
```

# Hypothesis tests
```{r}
#### HYPOTHESIS TESTS ####
#### NOTE: wilcox.test can't take gov_scale in factor form, since it needs to
#### add the ranks together to perform the hypothesis test. That column
#### is factored below for the visualizations, since the likert package
#### takes factors as inputs.

## HT 1): tested positive vs governor approval
# Subset q3_clean to create two vectors, since Wilcox.test two numeric vectors x, y as input. 
samples_positive_covid_test <- subset(q3_clean, covid_test_positive == 1)
samples_negative_covid_test <- subset(q3_clean, covid_test_positive == 2)

# paired=FALSE sets wilcox.test to perform a wilcoxon rank sum test. See ?wilcox.test for detail.
wilcox.test(x=samples_positive_covid_test$gov_scale, 
            y=samples_negative_covid_test$gov_scale,
            alternative="two.side",
            paired = FALSE, conf.level=0.95)

## HT 2): positive symptoms vs governor approval
# Same logic as with HT 1, except for covid_symptoms instead of positive_covid_test.
samples_covid_symptoms <- subset(q3_clean, covid_symptoms == 1)
samples_no_covid_symptoms <- subset(q3_clean, covid_symptoms == 2)

wilcox.test(x=samples_covid_symptoms$gov_scale, 
            y=samples_no_covid_symptoms$gov_scale,
            alternative="two.side",
            paired = FALSE, conf.level = 0.95)

```


# Practical Significance

```{r}
#### PRACTICAL SIGNIFICANCE ####
# REFERENCES/READING:
# https://statsandr.com/blog/correlation-coefficient-and-correlation-test-in-r/#correlation-coefficient
# https://stats.stackexchange.com/questions/3730/pearsons-or-spearmans-correlation-with-non-normal-data
# https://stats.stackexchange.com/questions/8071/how-to-choose-between-pearson-and-spearman-correlation
# https://rpubs.com/aaronsc32/spearman-rank-correlation
# https://statistics.laerd.com/statistical-guides/spearmans-rank-order-correlation-statistical-guide-2.php
# NOtation for spearman's rho: https://www.statisticssolutions.com/spearman-rank-correlation/


# Drop samples where covid test is below 0
q3_clean_covid_test <- subset(q3_clean, covid_test_positive > 0)

# This conducts the rank sum test and returns Spearman's Rho
cor.test(q3_clean_covid_test$covid_test_positive, 
                 q3_clean_covid_test$gov_scale, method="spearman",
                 exact=FALSE)

# This just returns Spearman's Rho
cor(q3_clean_covid_test$covid_test_positive, 
         q3_clean_covid_test$gov_scale, method="spearman")

# Drop samples where covid symptoms is below 0
q3_clean_covid_symptoms <- subset(q3_clean, covid_symptoms > 0)

cor.test(q3_clean_covid_symptoms$covid_symptoms, 
         q3_clean_covid_symptoms$gov_scale, method="spearman",
         exact=FALSE)

cor(q3_clean_covid_symptoms$covid_symptoms, 
    q3_clean_covid_symptoms$gov_scale, method="spearman")
```

